---
title: "Transportation Market Analysis"
author: "Bryan Blanc"
date: "8/7/2016"
output: html_document
---

This tutorial will guide you through using R to fetch, display, and export statistics and geographies typically useful in transportation (e.g., transit) market analyses. 

# Fetching Transit Supportive Population Statistics from ACS
American Community Survey data are used in a wide variety of applications by transportation planners. Until recently, much of this data required a fairly arduous querying, downloading, and cleaning process; the US Census Bureau's APIs and developers' query wrappers have made this task much easier. 

There is already a great R package for dealing with ACS data aptly named [`acs`](https://cran.r-project.org/web/packages/acs/index.html), which this vignette heavily relies upon.

Other additional information for utilizing the `acs` package can be found at the below links:

- [R Package Documentation](https://cran.r-project.org/web/packages/acs/acs.pdf)
- [A tutorial assembled by the package's creator](http://eglenn.scripts.mit.edu/citystate/wp-content/uploads/2013/06/wpid-working_with_acs_R3.pdf)
- [A great blog post covering some of the same topics as below](http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/)

## Setting up your workspace
If you have worked with R before, much of this should be familiar. There are a few initial things you should do to set up your workspace.

First you will want to load in some packages with functions you will need to follow the tutorial:
```{r,message=FALSE,warning=FALSE}
#Bryan's transportation data package. Includes some data you will need to work with this tutorial.
require(transportr)
#The workhorse package for querying ACS data
require(acs)
#Package for fetching census geography spatial data
require(tigris)
#Read/Write excel spreadsheets
require(XLConnect)
#The premier packages for data handling
require(dplyr)
require(tidyr)
#A great static plotting framework for R
require(ggplot2)
#Scales for plotting
require(scales)
#These packages includes the neccessary infrastructure for handling spatial data in R
require(sp)
require(rgdal)
require(rgeos)
require(taRifx.geo)
#This package has some handy functions for dealing with character strings
require(stringr)
#This package is used to create interactive maps
require(leaflet)
#These packages are used for a couple of HTML-related functions
require(htmltools)
require(htmlwidgets)
```

Then you need to set your working directory and load in your US Census Bureau API Key:
```{r,message=FALSE,warning=FALSE}
#Set working directory for input and output files
#Set your working directory and uncomment the below line
#setwd("/Users/bblanc/OneDrive/_transportr_development/_projects/censusData")
#Bryan's ACS API Key
api_key = "04976a4a378107d2fb53acdbde84d0aad121cb10"
#'Install' ACS API Key so you don't have to call it with every function
api.key.install(api_key)
```

## Creating an ACS query 
This tutorial uses data/queries used for the Hood River Transit Master Plan as a representative guide. 

To form an ACS query, you must first define four parameters of interest:

*Geography*
  : What geographical extent (e.g., the counties mentioned above) and resolution (e.g., census tract level) are you requesting data for?
  
*ACS Table*
  : What table contains the statistics you are interested in? This may take some exploring prior to your data query. You can search for tables using keywords with the `acs.lookup` function, or you can utilize the [American Fact Finder web site](http://factfinder.census.gov/faces/nav/jsf/pages/index.xhtml) to find the specific table you are interested in. 
  
*End Year*
  : ACS data are catalogued by the last year in the sample. What end year are you interested in?

*Span*
  : ACS data come in three flavors: 5-year, 3-year, and 1-year samples. 1-year samples are only available for areas where there was a large enough sample for statistics to be reliable. 5-year samples are available for all geographies, and are typically what is used in Nelson Nygaard market analyses.
  
Bryan developed the `transit_pops()` to take most of the work out of this query process, but you will still need to define your geography of interest. However, if you need statistics from different tables you may need to develop your own query; this requires some amount of customization. Herein, the `transit_pops()` function is relied upon to make this process as simple as possible. 

## Defining your geography
For the Hood River TMP, we are interested in block-group level statistics for several counties in Oregon and Washington:
  
- Oregon
    - Hood River County
    - Multnomah County 
    - Clackamas County 
    - Sherman County 
    - Wasco County
- Washington
    - Clark County
    - Klickitat County
    - Skamania County
    
You can define your geography using the `geo.make()` function from the `acs` package. A geography is made by defining several area and resolution parameters; these can be different depending on the geography of interest, but as the resolution increases the number of parameters neccessarily generally increases. Refer to the `geo.make()` help file for more information. Geographies in the `acs` package are grouped by state, so if you are requesting for more than one state you must add two queries together, as is done below. 
```{r}
#Here we are selecting all block groups, for all tracts, within the selected counties, within the selected states.
geog = geo.make(state="OR", county =c("Clackamas","Multnomah","Wasco","Hood River","Sherman"),
                tract="*", block.group ="*",check = TRUE)+
  geo.make(state="WA", county =c("Klickitat","Clark","Skamania"),
           tract="*", block.group ="*", check = TRUE)
```
## Querying the data
We're now ready to query data using the `transit_pops()` function. As long as your `year` is 2014 and your `span` is 5, all you need to worry about is the `geog` parameter. So plug in the geography we generated above. The querying takes a couple of minutes, depending on how big your geography is, but progress should be reported to the console. 
```{r Query ACS Data}
tpops_bg = transit_pops(geog=geog)
```
That's all there is to it for fetching the ACS data. 

# Joining ACS data to geography geometry
So now that you have the data, maybe you'd like to generate a map with it, but first we have to join the statistics to the relevant geography geometry. Thanks to the `tigris` package, and a helper function from `transportr`, this is very easy to do. Simply use the `geo_join()` function:
```{r Query ACS Data}
bg_data = geo_join_acs(tpops_bg,"block group")
```

# Fetching LODES data
The U.S. Census Bureau's Longitudinal Employee Household Dynamics (LEHD) division publishes LEHD Origin-Destination Employment Statistics annually. These statistics are useful for understanding commute flow patterns, employment density, residential density of employees, and other quanities of interest to transportation planners. 

To download LODES data, you will need to supply a *state* and *year* for your query, as well as what *file type* you are interested in. The following file types are available:

- Origin Destination (OD)
- Residential Area Characteristic (RAC)
- Workplace Area Characteristic (WAC)

More information on what variables are available in each of these file types is described in the LODES technical documentation (see latest technical doc: [http://lehd.ces.census.gov/data/](http://lehd.ces.census.gov/data/))

Using some helper functions from `transportr`, downloading LODES data is easy. First, you will want the geographic crosswalk so that you can understand which census blocks belong to which other geographies (e.g., block groups, tracts, places, counties). 
```{r,warning=FALSE,message=FALSE}
crosswalk = lodes_crosswalk("OR")
```

After fetching the crosswalk, you can then download the LODES files you're interested in. Some of these are large files, so the download may take a few minutes. After downloading, you will likely want to save out the files so there is no need to download them again. 

```{r,message=FALSE,warning=FALSE}
od_data = lodes_fetch("OR","OD",2014)
wac_data = lodes_fetch("OR","WAC",2014)
rac_data = lodes_fetch("OR","RAC",2014)

#Save out files as .csv or .rds objects
#write_csv(od_data,"od_data.csv")
#saveRDS(od_data,"od_data.rds")
```

# Aggregating and joining LODES data to geography geometry
Typically, we will want to utilize the LODES data at a higher geographic level than blocks. This will require aggregating the LODES data. Often, we will then want to join the data to the relevant geograpy geometry. 

```{r}
wac_bg = lodes_agg(wac_data,"WAC","block group",crosswalk)
wac_bg_geom = transportr::geo_join_lodes(wac_bg,type = "WAC","block group",crosswalk)

```



# Analyzing/Visualizing Data in R
Now that we've downloaded and joined both ACS and LODES data to our geographies, we can do some data analysis and visualizaton. 

## Non-spatial 
Even though we went through all the work of joining those geographies, we may want to do some statistical analysis or visualization ignorant of the geometric/spatial components of the data. 

## Spatial 
The most common application of this data is to generate demographic maps. 

### Leaflet maps

### Density calculation/mapping

# Exporting Data from R
Often, you will want to carry out data assembly in ArcGIS for export to Adobe Illustrator. To do this, we must first save our dataset as a shapefile. This is relatively easy:
```{r Save out ESRI Shapefile}
writeOGR(bg_data,"example","block_group_data",
         driver = "ESRI Shapefile",overwrite_layer = TRUE)
```


